Hydra Battles and Cie

### 2021-09-15 doc/chapter-powers.tex
corrections Chapter 10

Thus, any function able to compute more or less efficiently powers in a monoid will
give an algorithm for computing Fibonacci numbers. Unlike the \emph{ad-hoc} aforementioned proofs of~\cite{BC04}, the correctness of such an algorithm is a direct consequence
of the correctness of the used powering function.
Several examples will be presented in the rest of this document
(in sections~\vref{sect:fibonacci-pos-bpow}).

@@

We are now ready to define a new notation scope, in which the notation
\texttt{x * y} will be interpreted as an application of the function
\texttt{mult\_op}.

@@

The following instance of \texttt{EMonoid} describes the set of integers modulo
$m$, where $m$ is any integer greater than or equal to $2$.
For simplicity's sake, we represent such values using the \texttt{N} type,
and consider ``equivalence modulo \texttt{$m$}'' instead of equality.
Note that the law of associativity has been stated as Leibniz' equality.

@@

Therefore, we note that the term (\Verb|fun (x:A) =>x ^b 17|)  is
convertible, --- \emph{thus logically indistinguishable} ---, with a function that performs 16 multiplications.

@@

From these tests, we may infer that  representing exponentiation algorithms as \coq{}  functions hides
information about the real structure of the computations, particularly about  sharing intermediate computations.

Thus, we propose to define a data structure that makes explicit the sequence of multiplications that lead to the computation of $x^n$. For instance, the values of  
\texttt{x * x} and
\texttt{x * x * (x * x)}  are used
twice in the  computation of $x^{17}$ with the binary algorithm. This information should 
appear explicitly in the data structure chosen for representing exponentiation 
algorithms.

@@

\begin{quote}
For any item $i$ of $c$, there exists $j$ and $k$ in $c$, where
$i=j+k$, and $x^j$ and $x^k$ are already computed.

@@

In addition to this big computation time, this approach 
generates a huge proof term. Just try to execute the command 
``\texttt{Print C87\_ok}'' to get a measure of its size.
In order to understand this poor performance, let us consider an intermediate
subgoal of the previous proof generated after a sequence of unfoldings and simplifications. This goal is presented below.

\inputsnippets{Addition_Chains/WhySoSlow}

@@

\subsubsection{How does reflection work}
Let us consider again the subgoal on page~\pageref{fig:big-goal}, the conclusion of which has the form \texttt{|$a_1\,==\,a_2$|}, where \texttt{|$a_1$|} and
\texttt{|$a_2$|} are terms of  type \texttt{A}.
Instead of spending space and time in setoid rewritings, we would like to
normalize the terms \texttt{|$a_1$|} and \texttt{|$a_2$|} and verify that 
the associated normal forms are equal.

@@

\inputsnippets{Addition_Chains/MonoidExp}

Thus, the main steps of a correctness proof of a given chain, \emph{e.g.},
\texttt{C87} will be the following ones:
\begin{enumerate}
\item generate a subgoal as in page~\pageref{fig:big-goal},
\item express each term of the equivalence as the image of a term
     of type \texttt{Monoid\_Exp $A$},
\item normalize both terms and verify that their normal forms are equal.
\end{enumerate}

@@

We show how to \emph{prove} that  a given  chain $c$,
applied to any $a$, really computes $a^p$, where $p=\textrm{the\_exponent}\;c$.
Parametricity allows us to compare executions on any monoid $M$ 
with executions on \texttt{NatPlus}.
Let us consider the following mathematical relation 
$$\{(x,n)\in M\times\mathbb{N}\,|\, 0<n \wedge x=a^n\}$$

@@

This failure is due to a lack of an assumption that the continuation
\texttt{k} is \emph{proper} with respect to the equivalence \texttt{equiv}.
Thus, \coq{} is unable to infer from the equivalence 
\texttt{(a * a * a) == (a * (a * (a * E\_one)))} \linebreak that 
\texttt{k (a * a * a)} and \texttt{k (a * (a * (a * E\_one)))} are 
equivalent computations.

\subsubsection{Definition:} 
\index{coq}{Type classes}
\index{coq}{Type classes!Proper class}
A continuation \texttt{k:Fkont A} is \emph{proper}
if, whenever \linebreak[3] \texttt{x == y} holds, the computations \texttt{k x} and 
\texttt{k y} are equivalent.

@@

Let us show some manual correctness proofs of  small f-chains:

@@

\inputsnippets{Euclidean_Chains/Fplus}

@@

\pagebreak
\inputsnippets{Euclidean_Chains/Fplusb}

@@

\pagebreak
\inputsnippets{Euclidean_Chains/chainGen}
\inputsnippets{Euclidean_Chains/makeChain}
\inputsnippets{Euclidean_Chains/C87Dicho}

>>>>

Thus, any function able to compute more or less efficiently powers in a monoid will
give an algorithm for computing Fibonacci numbers. Unlike the \emph{ad-hoc} aforementioned proofs of~\cite{BC04}, the correctness of such an algorithm is a direct consequence
of the correctness of the used powering function.
Several examples will be presented in the rest of this document
(in Section~\vref{sect:fibonacci-pos-bpow}).

@@

We are now ready to define a new notation scope, in which the notation \linebreak 
\texttt{x * y} will be interpreted as an application of the function
\texttt{mult\_op}.

@@

The following instance of \texttt{EMonoid} describes the set of integers modulo
$m$, where $m$ is any integer greater than or equal to $2$.
For simplicity's sake, we represent such values using the \texttt{N} type,
and consider ``equivalence modulo \texttt{$m$}'' instead of equality.

@@

Therefore, we note that the term (\Verb|fun (x:A) =>x ^b 17|)  is
convertible --- \emph{thus logically indistinguishable} --- with a function that performs 16 multiplications.

@@

From these tests, we may infer that  representing exponentiation algorithms as plain arithmetic  functions hides
information about the real structure of the computations, particularly about  sharing intermediate computations.

Thus, we propose to define a data structure that makes explicit the sequence of multiplications that lead to the computation of $x^n$. For instance, the values of  
\texttt{x * x} and
\texttt{x * x * (x * x)}  are used
twice in the  computation of $x^{17}$ with the binary algorithm. This information should 
appear explicitly in the data structure chosen for representing
and comparing exponentiation 
algorithms.

@@

\begin{quote}
For any item $i$ of $c$ (but the first one), there exists $j$ and $k$ in $c$, where
$i=j+k$, and $x^j$ and $x^k$ are already computed.

@@

In addition to this big computation time, this approach 
generates a huge proof term. Just try to execute the command 
``\texttt{Print C87\_ok}'' to get a measure of its size.
In order to understand this poor performance, let us consider an intermediate
subgoal of the previous proof generated after a sequence of unfoldings and simplifications. This goal is presented below.
\begin{figure}[h]
  \centering
\inputsnippets{Addition_Chains/WhySoSlow}  
  \caption{A big goal}
  \label{fig:big-goal}
\end{figure}

@@

\subsubsection{How does reflection work}
Let us consider again the subgoal on Fig.~\vref{fig:big-goal}, the conclusion of which has the form \texttt{|$a_1\,==\,a_2$|}, where \texttt{|$a_1$|} and
\texttt{|$a_2$|} are terms of  type \texttt{A}.
Instead of spending space and time in setoid rewritings, we would like to
normalize the terms \texttt{|$a_1$|} and \texttt{|$a_2$|} and verify that 
the associated normal forms are equal.

@@

\pagebreak
\inputsnippets{Addition_Chains/MonoidExp}

Thus, the main steps of a correctness proof of a given chain, \emph{e.g.},
\texttt{C87} will be the following ones:
\begin{enumerate}
\item generate a subgoal as in Fig.~\vref{fig:big-goal},
\item express each term of the equivalence as the image of a term
     of type \texttt{Monoid\_Exp $A$},
\item normalize both terms and verify that their normal forms are equal.
\end{enumerate}

@@

We show how to \emph{prove} that  a given  chain $c$,
applied to any $a$, really computes $a^p$, where $p=\textrm{the\_exponent}\;c$.
Parametricity allows us to compare executions on any monoid $M$ 
with executions on \texttt{NatPlus}.
Let us consider the  mathematical relation 
$\{(x,n)\in M\times\mathbb{N}\,|\, 0<n \wedge x=a^n\}$.

@@

This failure is due to a lack of an assumption that the continuation
\texttt{k} is \emph{proper} with respect to the equivalence \texttt{equiv}.
Thus, \coq{} is unable to infer from the equivalence 
\texttt{(a * a * a) == (a * (a * (a * E\_one)))} \linebreak that 
(\texttt{k (a * a * a)}) and (\texttt{k (a * (a * (a * E\_one)))}) are 
equivalent computations.

\subsubsection{Definition:} 
\index{coq}{Type classes}
\index{coq}{Type classes!Proper class}
A continuation \texttt{k:Fkont A} is \emph{proper}
if, whenever \linebreak[3] \texttt{x == y} holds, the computations (\texttt{k x}) and 
(\texttt{k y}) are equivalent.

@@

Let us show manual correctness proofs of some small f-chains:

@@

\pagebreak
\inputsnippets{Euclidean_Chains/Fplus}

@@

\inputsnippets{Euclidean_Chains/Fplusb}

@@

\inputsnippets{Euclidean_Chains/chainGen}
\inputsnippets{Euclidean_Chains/makeChain}
\inputsnippets{Euclidean_Chains/makeChainCorrect}
\inputsnippets{Euclidean_Chains/C87Dicho}
